### 2021-12-10 manuscript.tex
Adjust formatting [ci skip]
This build is based on
https://github.com/lubianat/quali_phd/commit/bd9740705b22bf1d745d859aebf5f1b9fc80b3e1.

This commit was created by the following CI build and job:
https://github.com/lubianat/quali_phd/commit/bd9740705b22bf1d745d859aebf5f1b9fc80b3e1/checks
https://github.com/lubianat/quali_phd/actions/runs/1565041419

This manuscript
(\href{https://lubianat.github.io/quali_phd/v/cf629fe1f70f8f676f43e4ec04950c65a9674e6f/}{permalink})
was automatically generated
from \href{https://github.com/lubianat/quali_phd/tree/cf629fe1f70f8f676f43e4ec04950c65a9674e6f}{lubianat/quali\_phd@cf629fe}
on December 10, 2021.

@@

@@

\hypertarget{background}{%
\section{Background}\label{background}}

@@

Of note, even though a range of methods is available, most techniques and publications do not use standard identifiers for cell types.
This is in contradiction with the acknowledgement by the community of the advantages of using identifiers the ad using standard identifiers, such as those provided by the Cell Ontology. {[}\protect\hyperlink{ref-ii1dVqoy}{43}{]} {[}\protect\hyperlink{ref-tvnuubVW}{48}{]} {[}\protect\hyperlink{ref-J5X2Xu6M}{25}{]} {[}\protect\hyperlink{ref-agnqfdk6}{49}{]} {[}\protect\hyperlink{ref-SozKx0lF}{50}{]} {[}\protect\hyperlink{ref-15xnotFsw}{51}{]}.
Nevertheless, projects that use Cell Ontology identifiers for single-cell RNA-seq data are appearing {[}\protect\hyperlink{ref-qT8WxqjA}{52}{]}, including python and R packages (e.g.~Besca {[}\protect\hyperlink{ref-P1gDLuog}{53}{]}, OnClass {[}\protect\hyperlink{ref-sW6aNZJB}{54}{]} and ontoProc{[}\protect\hyperlink{ref-15YmDXALp}{55}{]}), data management projects and reference datasets, (e.g.~Tabula Muris {[}\protect\hyperlink{ref-tdlWsvTd}{56}/{]} and Tabula Sapiens {[}\protect\hyperlink{ref-FjUqVIOp}{57}{]} Azimuth map {[}\protect\hyperlink{ref-T7RxnUNw}{58}/{]} and HubMap's ASCT+B Tables {[}\protect\hyperlink{ref-1FghWzORJ}{59}{]}) and annotation platforms (e.g.~the Cell Annotation Platform {[}\protect\hyperlink{ref-yMavNVkS}{60}{]} and CellTypist {[}\protect\hyperlink{ref-123DiCgc8}{61}{]}.

@@

Regardless of ontology's sublanguage, it must be resolvable to an RDF-XML file.
RDF stands for Resource Description Framework, another W3C standard built around a graph-based data model {[}\protect\hyperlink{ref-djVboz3w}{72}/{]}.
Statements in RDF are triples consisting of 2 nodes (a subject and an object) and an edge (a predicate) connecting the nodes.
All nodes and edges are represented in RDFs by International Resource Identifiers (IRIs), and there are many ways to lay out those IRIs on a text file to represent triples.
One of those layouts is the RDF-XML syntax, inspired by the XML markup language.
Arguably, other syntaxes (interchangeable with RDF-XML) are easier to read for a human.
As an example of an RDF triple, here is how one would represent in the Turtle RDF Syntax, the notion that plasmacytoid dendritic cells are a type of dendritic cells:

@@

All the information is available on a user interface and programmatically.
Advanced users can download dumps in JSON and RDF dumps and acess the data via the MediaWiki API and a SPARQL endpoint. {[}\protect\hyperlink{ref-4wSLyAPr}{81}{]}
Several wrappers of such services are available in languages such as R {[}\protect\hyperlink{ref-1HghnypzG}{82}{]} and python {[}\protect\hyperlink{ref-VKWfrtR1}{83}/{]}.
The data scheme can be seen in \ref{fig:data} model, where each item is connected to a statement node via a property in the ``p:'' namespace, from which references and qualifiers are accessible.
To facilitate primary usage, the namespace ``wdt:'' connects items to values directly, simplifying, for example, the writing of SPARQL queries.

@@

Information on Wikidata is released under a CC0 license, which enables full reuse of the data. {[}\protect\hyperlink{ref-tQIJPG4k}{84}{]}
One of the major points of access and reuse of the information is the Wikidata Query Service {[}\protect\hyperlink{ref-2wDsXBwd}{85}/{]}, a core resource of the community which enables live querying in the SPARQL language. {[}\protect\hyperlink{ref-t66jYhqe}{86}{]}
A number of services make use of embedded queries from the Wikidata Query Service {[}\protect\hyperlink{ref-2wDsXBwd}{85}/{]} to create interactive, live dashboards like Scholia {[}\protect\hyperlink{ref-17P4zruFZ}{87}/{]} and the SARS-CoV-2 Query Book {[}\protect\hyperlink{ref-guMp3q1V}{88}/{]}

@@

Due to its privileged position inside the linked data ecosystem and its ease of writing and query, Wikidata has been growing as a hub for interoperable data for the life sciences community. {[}\protect\hyperlink{ref-3GqlN9Dk}{94}{]} {[}\protect\hyperlink{ref-Ym2HoRLl}{95}{]}
Even though Wikidata was created in 2013, the demand for a community-cured life sciences knowledge graph is apparent at least since 2008 {[}\protect\hyperlink{ref-179mXrGIk}{96}{]} {[}\protect\hyperlink{ref-TItj7EYf}{97}{]}
The Wikidata-like project proposed was eventually discontinued, an example of the challenge of maintaining independent biomedical databases. {[}\protect\hyperlink{ref-P2QdZvGc}{98}{]}
As Wikidata has a very large community, has stable funding and is at the core of modern technologies, like the Google Knowledge Graph {[}\protect\hyperlink{ref-xLpRePoh}{76}{]} and Amazon's Alexa, {[}\protect\hyperlink{ref-bO7BekzO}{99}/{]} it is virtually guaranteed that data in Wikidata will remain accessible for a long time, regardless of local funding schemes.

@@

Other projects outside the Gene Wiki initiative also started using Wikidata as a platform for knowledge integration.
A list of several projects that use Wikidata as part of their service to their community is given in table 1.
There is a movement exploring how Wikidata can be employed to advance Computational Biology and how it can be integrated to the current publication status quo. {[}\protect\hyperlink{ref-1GijvCQtv}{105}{]}
In that direction, Wikidata is being developed as a platform for scholarly linked open data, mainly via the Scholia platform {[}\protect\hyperlink{ref-hxzL9pmm}{106}{]} {[}\protect\hyperlink{ref-PdG2G4Bp}{107}{]},(\url{https://scholia.toolforge.org/}) which provides profiles of pre-templated SPARQL queries for entities like particular authors and articles (e.g.~Scholia profile on Prof.~Helder Nakaya available at \url{https://scholia.toolforge.org/author/Q42614737}).

\hypertarget{table-1}{%
\section{\texorpdfstring{\href{https://docs.google.com/spreadsheets/d/1loPE0bfJsih4eLrwNGvQf2c-9_I8z-GEvV21496o8h8/edit\#gid=0}{Table 1}}{Table 1}}\label{table-1}}

During the COVID-19 pandemic, Wikidata has spawned as a hotspot for modelling information about the virus and the pandemic in real-time. {[}\protect\hyperlink{ref-zH3KDqhp}{108}{]} {[}\protect\hyperlink{ref-wikidata:99196713}{\textbf{wikidata:99196713?}}{]}
The general scope of the database allowed representation in a shared system of molecular, epidemiologic and socio-economic aspects of the pandemic. {[}\protect\hyperlink{ref-zH3KDqhp}{108}{]}{[}\protect\hyperlink{ref-mPoPwN77}{109}{]}
Information curated in Wikidata was immediately available, feeding live dashboards and other applications based on SPARQL queries. {[}\protect\hyperlink{ref-3YCzr3jU}{110}{]} {[}\protect\hyperlink{ref-SkOeuFRd}{111}{]} {[}\protect\hyperlink{ref-AOVCR8mW}{112}{]}
Additionally, as the information presented on Wikidata is multilingual and collaboratively edited, it presented itself as a resource for constructing structured vocabularies in non-English languages. {[}\protect\hyperlink{ref-19wiU1LCA}{113}{]}

In addition to its value as a structured database, Wikidata is tightly connected to Wikipedia.
The gene identifiers in the context of Gene Wiki {[}\protect\hyperlink{ref-2ZhxC0dg}{102}{]} are now fed to Wikipedias across languages, benefitting users directly.
Additionally, gene expression information from the Bgee database {[}\protect\hyperlink{ref-10EWNsZCz}{114}{]} was added to Wikidata and connected to Wikipedia, which led to a sizeable increase of access to the Bgee database.
Currently, Wikipedia is one of the top 3 sources from which people access Bgee (personal communication with Tarcisio Farias), thus leading to direct recognition for integrated bases.
More generally, the connections of Wikidata and Wikipedia make it unique in the power of flowing knowledge back to human-accessed interfaces.
In the words of Matthias Samwald {[}\protect\hyperlink{ref-864Pcm4C}{115}{]} and colleagues, ``Wikidata could emerge as a community-backed and highly visible structured knowledge base of medical and biological information, bringing concepts and methodologies such as controlled taxonomies, Semantic Web / semantic technologies and ontologies into mainstream use.''

@@

\hypertarget{objectives}{%
\section{Objectives}\label{objectives}}

@@

\hypertarget{methodology}{%
\section{Methodology}\label{methodology}}

This project's methodology resembles practical research-action practices {[}\protect\hyperlink{ref-11UV98PHT}{116}{]}.
Our goal of improving the interoperability of cell-type data implies a combination of action and research.
Action in the form of active contributions to ontologies and knowledge graphs by getting involved and contributing to ongoing projects in the Human Cell Atlas and knowledge management context.
Research in 3 forms:
- Philosophical investigation on the nature of knowledge representations of cell types, both in formal logic settings and in current academic practice
- Applied investigations of database integration and data quality in the context of Wikidata and biomedical ontologies
- Data-driven biomedical research targeted at hypothesis generation and literature-based discovery using knowledge at the level of cell-type

@@

As part of the research-action process, I have joined the Cell Ontology working group.
I participate in the monthly meetings and sporadic workshops, learning and contributing to the discussions.
Additionally, I contribute to the ontology development, actively engaging in the Cell Ontology GitHub repository (\url{https://github.com/obophenotype/cell-ontology}) and contributing with new terms and assertions.
I edit the ontology with the software for ontology editing Protégé v. 5.5.0 (\url{https://protege.stanford.edu/}).

@@

\hypertarget{preliminary-results}{%
\section{Preliminary Results}\label{preliminary-results}}

@@

We targeted the question: which cell type definition allows crafting coherent biological statements?
The goal was not to say what cell types \emph{are}, but what they can be for a consistent representation on an ontology or a knowledge graph, like Wikidata.
We avoided the dissection of the differences between persistent classes of cells (often called ``cell types'') or the transient, fugacious classes of cells (often called ``cell states'') (see ``Definition of cell identity'' section in {[}\protect\hyperlink{ref-pFijpXkl}{117}{]} for an example).
Even though such a distinction is essential for theoretical research, it is not required to represent biomedical experiments formally.

In a preprint derived from this PhD project {[}\protect\hyperlink{ref-uCE45Jh8}{118}{]}, we proposed naming conventions for different cell types classes.
Much of the literature mixes cell types in one species (e.g., when dealing with a cell type as an evolutionary unit) or multiple species (e.g., in the Cell Ontology).
It is helpful to distil these different concepts into names. Given the importance of the species' concept in biological classification {[}\protect\hyperlink{ref-AHGRV6KQ}{119}{]}, we derive a species-centric view on naming classes of cell types.
The four classes (Figure \ref{fig:classes}) we propose are as follows:

@@

Refining the different concepts around the notion of ``cell type'' is important but will require decades before a reasonable consensus. Here we adopt a liberal view of cell type, defining, for our purposes, a cell type as any class of cells described by a domain expert with evidence of the reality of its instances.
The requirement of evidence of existence in the material world is based on the principle of instantiation of ontological realism {[}\protect\hyperlink{ref-5ZFRyTy5}{120}{]}. Barry Smith and Werner Ceuster state ``A term should be included in a reference ontology only if there is experimental evidence that instances to which that term refers exist in reality. (`Exists' here should be understood in a tenseless sense in order to accommodate, for example, universals pertaining to extinct species as well as universals such as swarm or hurricane which are instantiated only intermittently.)''
Thus, in this work, one minimum requirement for a cell type to be catalogued is a public description of the class, with evidence for the existence of instances of the class in reality.

By ``class'', we mean an abstract entity in a sense intended by the multilevel theory (MLT) of conceptual modelling {[}\protect\hyperlink{ref-eP47FILn}{121}{]}
Figure \ref{fig:multilevel} displays a simplified version of MLT adopted throughout this project.
In this framework, real-world entities are materializations of \emph{individuals}.
\emph{Individuals} are theoretical constructs that are (1) thought to exist or have existed, as per the principle of instantiation, and (2) refer to only one (01) material entity at any point in time.
For example, Wikidata has entries for people, e.g.~``Helder Nakaya (\href{https://www.wikidata.org/wiki/Q42614737}{Q42614737})'' and ``Charles Darwin (\href{https://www.wikidata.org/wiki/Q42614737}{Q1035})'' are considered \emph{individuals} by Multi-Level Theory.
Other examples of \emph{individuals} include ``Albert Einstein's brain (\href{https://www.wikidata.org/wiki/Q2464312}{Q2464312})'' and the ``Christ the Redeemer statue (\href{https://www.wikidata.org/wiki/Q79961}{Q79961})''.

@@

For the practical purpose adopted here, we avoid the dissection of the differences between persistent classes of cells (often called ``cell types'') or the transient, fugacious classes of cells (often called ``cell states'') (see ``Definition of cell identity'' section in {[}\protect\hyperlink{ref-pFijpXkl}{117}{]} for an example). We also consider only the cell as it was observed in an experiment, not necessarily the future conditions of any cell (i.e., the ``cell fate''). {[}\protect\hyperlink{ref-OjxQDZpx}{31}{]}\\
Even though such a distinction is an important topic for theoretical research, it is outside the initial scope of this work.

@@

We also note that we made a judgment call to use the term ``cell type'' to emphasize the focus on types as classes (or ``kinds'') in contrast to real-world objects.
The term ``cell class'' is also used in the literature and is a suitable synonym for our notion of cell type.
We opted to frame our work around the term ``cell type'' due to its historical usage and familiarity for the life sciences community.
Other related terms a ``cell set,'' ``cell population,'' and ``cell cluster,'' can also reminisce of a specific, countable group of cells, frequently from the same experiment.
The term ``cell identity'' has also been suggested for avoiding the cell type/cell state dilemma {[}\protect\hyperlink{ref-WKbly37M}{42}{]}, but we avoid it to emphasize a nominalistic perspective (in the Popperian sense{[}\protect\hyperlink{ref-1C6LI68h6}{122}{]}).
In doing so, we reinforce the intent to represent the cell types \_reported to exist\_instead of stating bluntly which cell types \emph{exist} or are \emph{essential} for human beings.

@@

Making the Human Cell Atlas more useful via Wikidata also includes the connection of related databases.
PanglaoDB {[}\protect\hyperlink{ref-M0lqRYJb}{123}{]} {[}\protect\hyperlink{ref-T3PlZ3Vh}{124}{]} is a publically-available database that contains data and metadata on hundreds of single-cell RNA sequencing experiments.
It provides extensive information on cell types, genes, tissues, and cell type markers, obtained via automatic and manual methods.
It also displays a rich web user interface for easy data acquisition, including database dumps for bulk downloads.

As of 8 December June 2021, the article describing PanglaoDB had been cited 230 times.
Despite its use by the community, the database is on a 3-star category for Linked Open Data {[}\protect\hyperlink{ref-OGXYtGZ8}{125}{]} as it does not use the open semantic standards from W3C (RDF and SPARQL) needed for a 4-star rank, neither the links to external data via standard identifiers that make datasets 5-star.
Improving the data format toward W3C's gold standards is a valuable step in making biological knowledge FAIR (Findable, Accessible, Interoperable, and Reusable).
Thus, we aimed to provide a case study of making the core information of PandlaoDB available in a 5-star Linked Open Data Format while improving the modelling of the necessary concepts on Wikidata.

As of August 2020, Wikidata had 264 items categorized as a ``cell type'', considerably less than in specialized cell catalogues, which count over two thousand cell types {[}\protect\hyperlink{ref-agnqfdk6}{49},\protect\hyperlink{ref-4AEy2xhQ}{126}{]}.
Strikingly, there were also 23 items categorized as instances of ``cell (Q7868)''.
This classification is imprecise, as an instance of cell would be an individual named cell from a single named individual, an example of conceptual disarray that often occurs on Wikidata. {[}\protect\hyperlink{ref-imSZTi52}{127}{]}

@@

The source code for Wikidata Bib is available at https://github.com/lubianat/wikidata\_bib.

>>>>

This manuscript
(\href{https://lubianat.github.io/quali_phd/v/bd9740705b22bf1d745d859aebf5f1b9fc80b3e1/}{permalink})
was automatically generated
from \href{https://github.com/lubianat/quali_phd/tree/bd9740705b22bf1d745d859aebf5f1b9fc80b3e1}{lubianat/quali\_phd@bd97407}
on December 10, 2021.

@@

\hypertarget{preface}{%
\subsection{Preface}\label{preface}}

@@

\hypertarget{background}{%
\subsection{Background}\label{background}}

@@

Of note, even though a range of methods is available, most techniques and publications do not use standard identifiers for cell types.
This is in contradiction with the acknowledgement by the community of the advantages of using identifiers the ad using standard identifiers, such as those provided by the Cell Ontology. {[}\protect\hyperlink{ref-ii1dVqoy}{43}{]} {[}\protect\hyperlink{ref-tvnuubVW}{48}{]} {[}\protect\hyperlink{ref-J5X2Xu6M}{25}{]} {[}\protect\hyperlink{ref-agnqfdk6}{49}{]} {[}\protect\hyperlink{ref-SozKx0lF}{50}{]} {[}\protect\hyperlink{ref-15xnotFsw}{51}{]}.
Nevertheless, projects that use Cell Ontology identifiers for single-cell RNA-seq data are appearing {[}\protect\hyperlink{ref-qT8WxqjA}{52}{]}, including python and R packages (e.g.~Besca {[}\protect\hyperlink{ref-P1gDLuog}{53}{]}, OnClass {[}\protect\hyperlink{ref-sW6aNZJB}{54}{]} and ontoProc{[}\protect\hyperlink{ref-15YmDXALp}{55}{]}), data management projects and reference datasets, (e.g.~Tabula Muris {[}\protect\hyperlink{ref-tdlWsvTd}{56}{]} and Tabula Sapiens {[}\protect\hyperlink{ref-FjUqVIOp}{57}{]} Azimuth map {[}\protect\hyperlink{ref-T7RxnUNw}{58}{]} and HubMap's ASCT+B Tables {[}\protect\hyperlink{ref-1FghWzORJ}{59}{]}) and annotation platforms (e.g.~the Cell Annotation Platform {[}\protect\hyperlink{ref-yMavNVkS}{60}{]} and CellTypist {[}\protect\hyperlink{ref-123DiCgc8}{61}{]}.

@@

Regardless of ontology's sublanguage, it must be resolvable to an RDF-XML file.
RDF stands for Resource Description Framework, another W3C standard built around a graph-based data model {[}\protect\hyperlink{ref-djVboz3w}{72}{]}.
Statements in RDF are triples consisting of 2 nodes (a subject and an object) and an edge (a predicate) connecting the nodes.
All nodes and edges are represented in RDFs by International Resource Identifiers (IRIs), and there are many ways to lay out those IRIs on a text file to represent triples.
One of those layouts is the RDF-XML syntax, inspired by the XML markup language.
Arguably, other syntaxes (interchangeable with RDF-XML) are easier to read for a human.
As an example of an RDF triple, here is how one would represent in the Turtle RDF Syntax, the notion that plasmacytoid dendritic cells are a type of dendritic cells:

@@

All the information is available on a user interface and programmatically.
Advanced users can download dumps in JSON and RDF dumps and acess the data via the MediaWiki API and a SPARQL endpoint. {[}\protect\hyperlink{ref-4wSLyAPr}{81}{]}
Several wrappers of such services are available in languages such as R {[}\protect\hyperlink{ref-1HghnypzG}{82}{]} and python {[}\protect\hyperlink{ref-VKWfrtR1}{83}{]}.
The data scheme can be seen in \ref{fig:data} model, where each item is connected to a statement node via a property in the ``p:'' namespace, from which references and qualifiers are accessible.
To facilitate primary usage, the namespace ``wdt:'' connects items to values directly, simplifying, for example, the writing of SPARQL queries.

@@

Information on Wikidata is released under a CC0 license, which enables full reuse of the data. {[}\protect\hyperlink{ref-tQIJPG4k}{84}{]}
One of the major points of access and reuse of the information is the Wikidata Query Service {[}\protect\hyperlink{ref-2wDsXBwd}{85}{]}, a core resource of the community which enables live querying in the SPARQL language. {[}\protect\hyperlink{ref-t66jYhqe}{86}{]}
A number of services make use of embedded queries from the Wikidata Query Service {[}\protect\hyperlink{ref-2wDsXBwd}{85}{]} to create interactive, live dashboards like Scholia {[}\protect\hyperlink{ref-17P4zruFZ}{87}{]} and the SARS-CoV-2 Query Book {[}\protect\hyperlink{ref-guMp3q1V}{88}{]}

@@

Due to its privileged position inside the linked data ecosystem and its ease of writing and query, Wikidata has been growing as a hub for interoperable data for the life sciences community. {[}\protect\hyperlink{ref-3GqlN9Dk}{94}{]} {[}\protect\hyperlink{ref-Ym2HoRLl}{95}{]}
Even though Wikidata was created in 2013, the demand for a community-cured life sciences knowledge graph is apparent at least since 2008 {[}\protect\hyperlink{ref-179mXrGIk}{96}{]} {[}\protect\hyperlink{ref-TItj7EYf}{97}{]}
The Wikidata-like project proposed was eventually discontinued, an example of the challenge of maintaining independent biomedical databases. {[}\protect\hyperlink{ref-P2QdZvGc}{98}{]}
As Wikidata has a very large community, has stable funding and is at the core of modern technologies, like the Google Knowledge Graph {[}\protect\hyperlink{ref-xLpRePoh}{76}{]} and Amazon's Alexa, {[}\protect\hyperlink{ref-bO7BekzO}{99}{]} it is virtually guaranteed that data in Wikidata will remain accessible for a long time, regardless of local funding schemes.

@@

Other projects outside the Gene Wiki initiative also started using Wikidata as a platform for knowledge integration.
A list of several projects that use Wikidata as part of their service to their community is given in table \ref{tbl:wikidata_databases1}.
There is a movement exploring how Wikidata can be employed to advance Computational Biology and how it can be integrated to the current publication status quo. {[}\protect\hyperlink{ref-1GijvCQtv}{105}{]}
In that direction, Wikidata is being developed as a platform for scholarly linked open data, mainly via the Scholia platform {[}\protect\hyperlink{ref-hxzL9pmm}{106}{]} {[}\protect\hyperlink{ref-PdG2G4Bp}{107}{]},(\url{https://scholia.toolforge.org/}) which provides profiles of pre-templated SPARQL queries for entities like particular authors and articles (e.g.~Scholia profile on Prof.~Helder Nakaya available at \url{https://scholia.toolforge.org/author/Q42614737}).

\begin{longtable}[]{@{}
  >{\raggedright\arraybackslash}p{(\columnwidth - 4\tabcolsep) * \real{0.1556}}
  >{\raggedright\arraybackslash}p{(\columnwidth - 4\tabcolsep) * \real{0.4000}}
  >{\raggedright\arraybackslash}p{(\columnwidth - 4\tabcolsep) * \real{0.4444}}@{}}
\caption{Samples of databases reconciled partially or totally to Wikidata
\label{tbl:wikidata_databases}}\tabularnewline
\toprule
\begin{minipage}[b]{\linewidth}\raggedright
name
\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedright
project website
\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedright
citation
\end{minipage} \\
\midrule
\endfirsthead
\toprule
\begin{minipage}[b]{\linewidth}\raggedright
name
\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedright
project website
\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedright
citation
\end{minipage} \\
\midrule
\endhead
LOTUS & \url{https://lotus.naturalproducts.net/} & {[}\protect\hyperlink{ref-P1EnNa7Q}{108}{]} \\
GeneDB & \url{https://www.genedb.org/} & {[}\protect\hyperlink{ref-gqY9YdmH}{109}{]} \\
Cellosaurus & \url{https://web.expasy.org/cellosaurus/} & {[}\protect\hyperlink{ref-HUr1iigG}{110}{]} \\
Complex Portal & \url{https://www.ebi.ac.uk/complexportal/} & {[}\protect\hyperlink{ref-k4BJEMHE}{111}{]} \\
WikiPathways & \url{https://www.wikipathways.org/} & {[}\protect\hyperlink{ref-pzgOjFLZ}{112}{]} \\
Reactome & \url{https://reactome.org/} & {[}\protect\hyperlink{ref-1Fqp7ppqH}{113}{]} \\
CIViC & \url{http://www.civicdb.org} & {[}\protect\hyperlink{ref-HUodcdz1}{114}{]} \\
PubChem & \textless https://pubchem.ncbi.nlm.nih.gov/ \textgreater{} & {[}\protect\hyperlink{ref-HUodcdz1}{114}{]} \\
Human Disease Ontology & \url{https://www.ebi.ac.uk/ols/ontologies/doid} & {[}\protect\hyperlink{ref-HUodcdz1}{114}{]} \\
\bottomrule
\end{longtable}

During the COVID-19 pandemic, Wikidata has spawned as a hotspot for modelling information about the virus and the pandemic in real-time. {[}\protect\hyperlink{ref-zH3KDqhp}{115}{]} {[}\protect\hyperlink{ref-wikidata:99196713}{\textbf{wikidata:99196713?}}{]}
The general scope of the database allowed representation in a shared system of molecular, epidemiologic and socio-economic aspects of the pandemic. {[}\protect\hyperlink{ref-zH3KDqhp}{115}{]}{[}\protect\hyperlink{ref-mPoPwN77}{116}{]}
Information curated in Wikidata was immediately available, feeding live dashboards and other applications based on SPARQL queries. {[}\protect\hyperlink{ref-3YCzr3jU}{117}{]} {[}\protect\hyperlink{ref-SkOeuFRd}{118}{]} {[}\protect\hyperlink{ref-AOVCR8mW}{119}{]}
Additionally, as the information presented on Wikidata is multilingual and collaboratively edited, it presented itself as a resource for constructing structured vocabularies in non-English languages. {[}\protect\hyperlink{ref-19wiU1LCA}{120}{]}

In addition to its value as a structured database, Wikidata is tightly connected to Wikipedia.
The gene identifiers in the context of Gene Wiki {[}\protect\hyperlink{ref-2ZhxC0dg}{102}{]} are now fed to Wikipedias across languages, benefitting users directly.
Additionally, gene expression information from the Bgee database {[}\protect\hyperlink{ref-10EWNsZCz}{121}{]} was added to Wikidata and connected to Wikipedia, which led to a sizeable increase of access to the Bgee database.
Currently, Wikipedia is one of the top 3 sources from which people access Bgee (personal communication with Tarcisio Farias), thus leading to direct recognition for integrated bases.
More generally, the connections of Wikidata and Wikipedia make it unique in the power of flowing knowledge back to human-accessed interfaces.
In the words of Matthias Samwald {[}\protect\hyperlink{ref-864Pcm4C}{122}{]} and colleagues, ``Wikidata could emerge as a community-backed and highly visible structured knowledge base of medical and biological information, bringing concepts and methodologies such as controlled taxonomies, Semantic Web / semantic technologies and ontologies into mainstream use.''

@@

\hypertarget{objectives}{%
\subsection{Objectives}\label{objectives}}

@@

\hypertarget{methodology}{%
\subsection{Methodology}\label{methodology}}

This project's methodology resembles practical research-action practices {[}\protect\hyperlink{ref-11UV98PHT}{123}{]}.
Our goal of improving the interoperability of cell-type data implies a combination of action and research.
Action in the form of active contributions to ontologies and knowledge graphs by getting involved and contributing to ongoing projects in the Human Cell Atlas and knowledge management context.
Research in 3 forms:
- Philosophical investigation on the nature of knowledge representations of cell types, both in formal logic settings and in current academic practice
- Applied investigations of database integration and data quality in the context of Wikidata and biomedical ontologies
- Data-driven biomedical research targeted at hypothesis generation and literature-based discovery using knowledge at the level of cell-type

@@

As part of the research-action process, I have joined the Cell Ontology working group.
I participate in the monthly meetings and sporadic workshops, learning and contributing to the discussions.
Additionally, I contribute to the ontology development, actively engaging in the Cell Ontology GitHub repository (\url{https://github.com/obophenotype/cell-ontology}) and contributing with new terms and assertions.
I edit the ontology with the software for ontology editing Protégé v. 5.5.0 {[}\protect\hyperlink{ref-CHsAoEvr}{124}{]}.

@@

\hypertarget{preliminary-results}{%
\subsection{Preliminary results}\label{preliminary-results}}

@@

We targeted the question: which cell type definition allows crafting coherent biological statements?
The goal was not to say what cell types \emph{are}, but what they can be for a consistent representation on an ontology or a knowledge graph, like Wikidata.
We avoided the dissection of the differences between persistent classes of cells (often called ``cell types'') or the transient, fugacious classes of cells (often called ``cell states'') (see ``Definition of cell identity'' section in {[}\protect\hyperlink{ref-pFijpXkl}{125}{]} for an example).
Even though such a distinction is essential for theoretical research, it is not required to represent biomedical experiments formally.

In a preprint derived from this PhD project {[}\protect\hyperlink{ref-uCE45Jh8}{126}{]}, we proposed naming conventions for different cell types classes.
Much of the literature mixes cell types in one species (e.g., when dealing with a cell type as an evolutionary unit) or multiple species (e.g., in the Cell Ontology).
It is helpful to distil these different concepts into names. Given the importance of the species' concept in biological classification {[}\protect\hyperlink{ref-AHGRV6KQ}{127}{]}, we derive a species-centric view on naming classes of cell types.
The four classes (Figure \ref{fig:classes}) we propose are as follows:

@@

Refining the different concepts around the notion of ``cell type'' is important but will require decades before a reasonable consensus. Here we adopt a liberal view of cell type, defining, for our purposes, a cell type as any class of cells described by a domain expert with evidence of the reality of its instances.
The requirement of evidence of existence in the material world is based on the principle of instantiation of ontological realism {[}\protect\hyperlink{ref-5ZFRyTy5}{128}{]}. Barry Smith and Werner Ceuster state ``A term should be included in a reference ontology only if there is experimental evidence that instances to which that term refers exist in reality. (`Exists' here should be understood in a tenseless sense in order to accommodate, for example, universals pertaining to extinct species as well as universals such as swarm or hurricane which are instantiated only intermittently.)''
Thus, in this work, one minimum requirement for a cell type to be catalogued is a public description of the class, with evidence for the existence of instances of the class in reality.

By ``class'', we mean an abstract entity in a sense intended by the multilevel theory (MLT) of conceptual modelling {[}\protect\hyperlink{ref-eP47FILn}{129}{]}
Figure \ref{fig:multilevel} displays a simplified version of MLT adopted throughout this project.
In this framework, real-world entities are materializations of \emph{individuals}.
\emph{Individuals} are theoretical constructs that are (1) thought to exist or have existed, as per the principle of instantiation, and (2) refer to only one (01) material entity at any point in time.
For example, Wikidata has entries for people, e.g.~``Helder Nakaya (\href{https://www.wikidata.org/wiki/Q42614737}{Q42614737})'' and ``Charles Darwin (\href{https://www.wikidata.org/wiki/Q42614737}{Q1035})'' are considered \emph{individuals} by Multi-Level Theory.
Other examples of \emph{individuals} include ``Albert Einstein's brain (\href{https://www.wikidata.org/wiki/Q2464312}{Q2464312})'' and the ``Christ the Redeemer statue (\href{https://www.wikidata.org/wiki/Q79961}{Q79961})''.

@@

For the practical purpose adopted here, we avoid the dissection of the differences between persistent classes of cells (often called ``cell types'') or the transient, fugacious classes of cells (often called ``cell states'') (see ``Definition of cell identity'' section in {[}\protect\hyperlink{ref-pFijpXkl}{125}{]} for an example). We also consider only the cell as it was observed in an experiment, not necessarily the future conditions of any cell (i.e., the ``cell fate''). {[}\protect\hyperlink{ref-OjxQDZpx}{31}{]}\\
Even though such a distinction is an important topic for theoretical research, it is outside the initial scope of this work.

@@

We also note that we made a judgment call to use the term ``cell type'' to emphasize the focus on types as classes (or ``kinds'') in contrast to real-world objects.
The term ``cell class'' is also used in the literature and is a suitable synonym for our notion of cell type.
We opted to frame our work around the term ``cell type'' due to its historical usage and familiarity for the life sciences community.
Other related terms a ``cell set,'' ``cell population,'' and ``cell cluster,'' can also reminisce of a specific, countable group of cells, frequently from the same experiment.
The term ``cell identity'' has also been suggested for avoiding the cell type/cell state dilemma {[}\protect\hyperlink{ref-WKbly37M}{42}{]}, but we avoid it to emphasize a nominalistic perspective (in the Popperian sense{[}\protect\hyperlink{ref-1C6LI68h6}{130}{]}).
In doing so, we reinforce the intent to represent the cell types \_reported to exist\_instead of stating bluntly which cell types \emph{exist} or are \emph{essential} for human beings.

@@

Making the Human Cell Atlas more useful via Wikidata also includes the connection of related databases.
PanglaoDB {[}\protect\hyperlink{ref-M0lqRYJb}{131}{]} {[}\protect\hyperlink{ref-T3PlZ3Vh}{132}{]} is a publically-available database that contains data and metadata on hundreds of single-cell RNA sequencing experiments.
It provides extensive information on cell types, genes, tissues, and cell type markers, obtained via automatic and manual methods.
It also displays a rich web user interface for easy data acquisition, including database dumps for bulk downloads.

As of 8 December June 2021, the article describing PanglaoDB had been cited 230 times.
Despite its use by the community, the database is on a 3-star category for Linked Open Data {[}\protect\hyperlink{ref-OGXYtGZ8}{133}{]} as it does not use the open semantic standards from W3C (RDF and SPARQL) needed for a 4-star rank, neither the links to external data via standard identifiers that make datasets 5-star.
Improving the data format toward W3C's gold standards is a valuable step in making biological knowledge FAIR (Findable, Accessible, Interoperable, and Reusable).
Thus, we aimed to provide a case study of making the core information of PandlaoDB available in a 5-star Linked Open Data Format while improving the modelling of the necessary concepts on Wikidata.

As of August 2020, Wikidata had 264 items categorized as a ``cell type'', considerably less than in specialized cell catalogues, which count over two thousand cell types {[}\protect\hyperlink{ref-agnqfdk6}{49},\protect\hyperlink{ref-4AEy2xhQ}{134}{]}.
Strikingly, there were also 23 items categorized as instances of ``cell (Q7868)''.
This classification is imprecise, as an instance of cell would be an individual named cell from a single named individual, an example of conceptual disarray that often occurs on Wikidata. {[}\protect\hyperlink{ref-imSZTi52}{135}{]}

@@

The source code for Wikidata Bib is available at \url{https://github.com/lubianat/wikidata_bib}.
